In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-20 paper/main.tex
Updates from Overleaf

\subsection*{In-silico predictions of intracranial transport after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/figure1.png}
     \caption{
     \textbf{In-silico modelling of transport of the PVS, SAS and parenchyma.}
     A) Illustration of the model geometry including the CSF-filled spaces (ventricles and SAS) $\Omega_{\rm CSF}$, the parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and veins $\Lambda_{\rm vein}$, as well as their interfaces and boundaries; 
     B) 3D rendering of the computational geometry: the parenchyma, CSF (clipped, blue), arterial network (red), and venous network (blue); 
     C) illustration of model representation of the intrathecal tracer injection; 
     D) in-silico predictions of tracer concentration after 1, 6, 12, and 24 hours (opacity increasing linearly from 0 to 2 mmol/l); 
     E) sagittal, coronal and axial view of in-silico tracer concentrations overlayed on T1-weighted MR image after 1, 6, 12 and 24\,h (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); 
     F) average tracer concentration in each compartment over the first 24\,h; 
     G) total amount of tracer in each compartment over the first 24\,h.}
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--B). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

@@

Our simulation results highlight that reproducing clinically observed tracer timings requires flow drivers beyond CSF production and cardiac peristalsis. To illustrate, we compared our findings to tracer arrival times reported by Eide and Ringstad~\cite{eide2024functional}, who observed a delay of just 15.3 min between the M1 and M2 segments of the MCA. Our baseline model, accounting only for CSF production and cardiac-driven flow, predicted a physiologically unrealistic delay of over three hours. In sharp contrast, our 'high PVS flow' scenario, which incorporates the effects of low-frequency vasomotion, yielded a much more consistent delay of 24 minutes. Furthermore, after adjusting for the 13 min travel time from injection to the craniocervical junction~\cite{eide2024functional}, the 'high PVS flow' model's FTAs (84––108 min for M1) aligns well with the upper end of the reported clinical range (37.8 $\pm$ 47.0), particularly when considering potentially delayed transport due to the limited caudal extend of our PVS network.
Therefore, we interpret these results as strong indication that PVS flow driven by CSF production and cardiac peristalsis alone is insufficient to explain rapid perivascular transport observed in humans.

@@

% Limitations
In terms of limitations, our primary constraint modelling-wise is the
geometric representation of the vascular (and therefore also
perivascular) networks. Concretely, we account for no vessels below
the original MRI resolution limits, and thus only a part of the
cerebral surface vasculature and effectively not the cortical
vessels. In addition, and again due to lack of resolution, there is
uncertainty associated with the location of the vessel segments
relative to the CSF spaces and parenchyma, with some vessels
surrounded by both domains and some vessels completely entering the
parenchyma before returning to the SAS. Higher resolution MRI data
(T1w and e.g. time-of-flight, QSM or other modalities) in more
subjects would immediately allow for more detailed studies of this
aspect. As a result of the lack of an accurate representation of the
cortical and subcortical white matter vasculature, we have considered
a basic transport model within the parenchyma, modelling extracellular
diffusion alone. To compensate, we have focused on reporting detailed
predictions in the CSF spaces and surface PVS after not more than 24
hours. In particular, we note that, we observe tracer accumulation at
the perivascular network ends after 12 hours, similar to that
previously observed for microspheres~\cite{bedussi2018paravascular,
  mestre2018flow}, which here can be interpreted as an artifact of the
incomplete network creating a barrier for tracer movement. Previously,
by inferring transport parameters within the human brain from
glymphatic MRI data, we have found that molecular transport within the
parenchyma is well-represented by $\approx 3\times$ enhanced diffusion
augmented by local clearance e.g.~across the blood-brain barrier, or
by diffusion-convection with a complex pattern of spatially-varying
interstitial fluid flow on the order of 1-10 \textmu
m/min~\cite{vinje2023human}. We will incorporate such transport
mechanisms within the parenchyma in future work. Finally, we also note
that we simulate the intracranial distribution of a tracer
concentration appearing at the craniocervical junction rather than as
an intrathecal injection. We therefore overestimate the amount of
tracer entering into the intracranial spaces; however, due to the
linearity of the mathematical model, it is valid to directly interpret
the in-silico concentrations relative to the total amount of tracer.

>>>>

\subsection*{In-silico predictions of intracranial transport after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/figure1.png}
     \caption{
     \textbf{In-silico modelling of transport of the PVS, SAS and parenchyma.}
     A) Illustration of the model geometry including the CSF-filled spaces (ventricles and SAS) $\Omega_{\rm CSF}$, the parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and veins $\Lambda_{\rm vein}$, as well as their interfaces and boundaries; 
     B) 3D rendering of the computational geometry: the parenchyma, CSF (clipped, blue), arterial network (red), and venous network (blue); 
     C) illustration of the intrathecal tracer injection protocol;  
     D) in-silico predictions of tracer concentration after 1, 6, 12, and 24 hours (opacity increasing linearly from 0 to 2 mmol/l); 
     E) sagittal, coronal and axial view of in-silico tracer concentrations overlayed on T1-weighted MR image after 1, 6, 12 and 24\,h (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); 
     F) average tracer concentration in each compartment over the first 24\,h; 
     G) total amount of tracer in each compartment over the first 24\,h.}
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--B). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

@@

Our simulation results highlight that reproducing clinically observed tracer timings requires flow drivers beyond CSF production and cardiac peristalsis.
Similar to the CSF spaces, dispersion due to cardiac and respiratory pulsation has been hypothesized to enhance transport in the perivascular spaces. However, recent estimates indicate that the effect of dispersion does not exceed a factor of 2 compared to diffusion alone \cite{asgari2016glymphatic,sharp2019dispersion,bojarskaite2023sleep,asgari2016glymphatic,troyetsky2021dispersion}. This conclusion is supported by our own model, where a 100-fold increase in the PVS dispersion coefficient did not substantially accelerate transport (\cref{fig:pvs_b}H). 
To illustrate the need for a stronger advective driver, we compared our findings to tracer arrival times reported by Eide and Ringstad~\cite{eide2024functional}, who observed a delay of just 15.3 min between the M1 and M2 segments of the MCA. Our baseline model, accounting only for CSF production and cardiac-driven flow, predicted a physiologically unrealistic delay of over three hours. In sharp contrast, our 'high PVS flow' scenario, which incorporates the effects of low-frequency vasomotion, yielded a much more consistent delay of 24 minutes. Furthermore, after adjusting for the 13 min travel time from injection to the craniocervical junction~\cite{eide2024functional}, the 'high PVS flow' model's FTAs (84––108 min for M1) aligns well with the upper end of the reported clinical range (37.8 $\pm$ 47.0), particularly when considering potentially delayed transport due to the limited caudal extend of our PVS network.
Therefore, we interpret these results as strong indication that PVS flow driven by CSF production and cardiac peristalsis alone is insufficient to explain rapid perivascular transport observed in humans.

@@

% Limitations
In terms of limitations, our primary constraint modelling-wise is the
geometric representation of the vascular (and therefore also
perivascular) networks. Concretely, we account for no vessels below
the original MRI resolution limits, and thus only a part of the
cerebral surface vasculature and effectively not the cortical
vessels. In addition, and again due to lack of resolution, there is
uncertainty associated with the location of the vessel segments
relative to the CSF spaces and parenchyma, with some vessels
surrounded by both domains and some vessels completely entering the
parenchyma before returning to the SAS. Higher resolution MRI data
(T1w and e.g. time-of-flight, QSM or other modalities) in more
subjects would immediately allow for more detailed studies of this
aspect. As a result of the lack of an accurate representation of the
cortical and subcortical white matter vasculature, we have considered
a basic transport model within the parenchyma, modelling extracellular
diffusion alone. To compensate, we have focused on reporting detailed
predictions in the CSF spaces and surface PVS after not more than 24
hours. In particular, we note that, we observe tracer accumulation at
the perivascular network ends after 12 hours, similar to that
previously observed for microspheres~\cite{bedussi2018paravascular,
  mestre2018flow}, which here can be interpreted as an artifact of the
incomplete network creating a barrier for tracer movement. Previously,
by inferring transport parameters within the human brain from
glymphatic MRI data, we have found that molecular transport within the
parenchyma is well-represented by $\approx 3\times$ enhanced diffusion
augmented by local clearance e.g.~across the blood-brain barrier, or
by diffusion-convection with a complex pattern of spatially-varying
interstitial fluid flow on the order of 1-10 \textmu
m/min~\cite{vinje2023human}. We will incorporate such transport
mechanisms within the parenchyma in future work. Finally, we also note
that we simulate the intracranial distribution of a tracer
concentration appearing at the craniocervical junction rather than as
an intrathecal injection. We thus neglect absorption of both CSF and tracer in the spinal compartment, and therefore overestimate the amount of
tracer entering into the intracranial spaces; however, due to the
linearity of the mathematical model, it is valid to directly interpret
the in-silico concentrations relative to the total amount of tracer.
